\begin{abstract}

\Adtpp\ is an open-source tool that adds support for algebraic data types
(ADTs) to the C programming language.
ADTs allow more precise description of program types and more robust handling
of data structures than is directly supported by C.
ADT definitions and other
declarations are put in a file that is preprocessed by \adtpp\ 
to produce a C header (``.h'') file which can be included in C
source files.  The generated header file contains C type definitions, macros,
and inline functions that support type-safe construction, deconstruction,
and pattern matching of ADT values, while
avoiding unsafe operations such as casts, and avoiding the risk of errors such as
dereferencing NULL pointers and accessing inappropriate
fields of unions.  Values are represented efficiently, using techniques
from the implementation of declarative languages.  For many simple data
types, the representation is identical to a direct implementation in C,
with no loss of efficiency.  For more complex
types, the \adtpp\ representation is more efficient than common C
representations, while preserving type safety and convenience.
As an example, we present a new variation of 234-trees which is
very compact.
\Adtpp\ also
supports parametric polymorphism such as defining a type ``list of $t$'',
where $t$ can be any ADT, and generic functions such as length.  However,
compared to typical declarative languages, polymorphic code is somewhat
more verbose, due to our reliance on the limited type checking available
in C.

\noindent
\textbf{Keywords}: Algebraic data types, C language extension, generics,
polymorphism, safety, static checking, open-source
\end{abstract}
