\documentclass[times]{speauth}
\begin{document}
\newcommand{\adtpp}{\texttt{adtpp}}
\newcommand{\Adtpp}{\texttt{Adtpp}}

\runningheads{L.Naish, P. Schachte and A. MacNally}{Adtpp: ADTs for C}
\title{Adtpp: lightweight efficient safe polymorphic algebraic data
types for C\footnote{to appear in \textit{Software Practice and Experience}}}
\author{Lee Naish\affil{1}, Peter Schachte\affil{1} and
Aleck MacNally\affil{1}}
\address{\affilnum{1}Computing and Information Systems, Universtity of
Melbourne}
\corraddr{
E-mail: \{lee,schachte\}@unimelb.edu.au}
\begin{abstract}

\Adtpp\ is an open-source tool that adds support for algebraic data types
(ADTs) to the C programming language.  ADTs allow more precise description
of program types and more robust handling of data structures than is
directly supported by C.  ADT definitions and other declarations are put
in a file that is preprocessed by \adtpp\ to produce a C header (``.h'')
file which can be included in C source files.  The generated header file
contains C type definitions, macros, and inline functions that support
type-safe construction, deconstruction, and pattern matching of ADT
values, while avoiding unsafe operations such as casts, and avoiding
the risk of errors such as dereferencing NULL pointers and accessing
inappropriate fields of unions.  Values are represented efficiently,
using techniques from the implementation of declarative languages.
For many simple data types, the memory representation is identical to a direct
implementation in C, with no loss of efficiency.  For more complex
types, the \adtpp\ representation is more efficient than common C
representations, while preserving type safety and convenience.  As an
example, we present a new variation of 234-trees which is very compact.
\Adtpp\ also supports parametric polymorphism such as defining a type
``list of $t$'', where $t$ can be any ADT, and generic functions
such as length.  However, polymorphic code is somewhat more verbose
than for typical declarative languages, due to our reliance on the
limited type checking available in C.

\end{abstract}
\keywords{Algebraic data types; C language extension; generics;
polymorphism; safety; static checking; open-source}
\maketitle

\input{1}

\bibliographystyle{wileyj}
\bibliography{all}
\end{document}

